---
title: "200414 Dplyr"
author: "Vegard Lysne"
date: "7 4 2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

# Select dataset
```{r include = F}
library(help = "datasets")
datasets

```

# Select variables/columns
1. Extract **vector** from tibble (pull)
2. **Subset** tibble (select)
3. select-helpers

## Extract vector
```{r}
manu <- mpg %>% 
  distinct(manufacturer) %>% 
  pull(manufacturer)

mpg %>% 
  pull(displ) %>% 
  range()

mpg[["displ"]]
```

## Subset/select variable
```{r}
mpg %>% 
  select(cyl, manufacturer, model)

mpg %>% 
  select_if(is.numeric)
```

## Select-helpers
- from tidyselect
```{r}
mpg %>% select(starts_with("c"))

mpg %>% select(ends_with("l"))

mpg %>% select(contains("isp"))

mpg %>% select(cyl, everything())

x <- c("manufacturer", "cyl", "model")

mpg %>% select(all_of(c("manufacturer", "cyl")))

mpg %>% select(starts_with("c"), everything())
```


# Create new or modify variables
- vectorised functions returns a vector of equal length
```{r}
mpg %>% 
  rename(newname = manufacturer)

mpg %>% select(newname = manufacturer, everything())

mpg %>% select(manufacturer, model, displ)

mpg %>% 
  mutate(cyl2 = cyl*2,
         cyl3 = cyl*3,
         cyl4 = cyl2*cyl3) %>%  select(cyl, cyl2, cyl3, cyl4, everything())

mpg %>% 
  mutate(newvar = 1, 
         anothernewvar = 2, 
         third = newvar + anothernewvar)


mpg %>% 
  mutate(displ = log(displ),
         displ = scale(displ))
```

# Select cases/rows
- filter rows if expression = TRUE

## Filter
- logical operators (==, !=, >, <)
- logical tests (is.na, !is.na, between(), etc)
```{r}
mpg %>% filter(cyl == 4)
mpg %>% filter(cyl != 4)
mpg %>% filter(cyl >= 4)
mpg %>% filter(displ > mean(displ))

mpg %>% filter(manufacturer == "audi")
mpg %>% filter(between(displ, 3, 5))
mpg %>% filter(str_detect(manufacturer, "au"))
```

## Filter on more conditions
- Boolean operators (&, |)
```{r}
mpg %>% 
  filter(manufacturer == "audi" & cyl == 4)

mpg %>% 
  filter(manufacturer == "audi" | cyl == 4 | displ > 5)

vect <- c("audi", "dodge")
mpg %>% 
  filter(manufacturer %in% c("audi", "dodge"))

mpg %>% 
  filter(!(manufacturer %in% c("audi", "dodge")))

mpg %>% filter(!is.na(cyl))
```

## Arrange by value
-desc() for decreasing order
```{r}
mpg %>% 
  arrange(year, cyl, desc(displ)) %>% 
  select(year, cyl, displ) %>%
  group_by(year, cyl) %>%
  slice(1)

mpg %>% 
  arrange(year, desc(cyl), manufacturer)
```



## Subset rows by position
```{r}
mpg %>% slice(1)
mpg %>% slice(1:5)
```

# Group dataset
```{r}
mpg %>% 
  group_by(manufacturer) %>% 
  str()

mpg %>% 
  group_by(manufacturer) %>% 
  slice(1)

mpg %>% 
  group_by(manufacturer, model) %>% 
  slice(1:2)
```

# Summarise
```{r}

mpg %>%
  group_by(year, cyl) %>% 
  summarise(quant = quantile(displ, probs = c(0.25, 0.5, 0.75)),
            probs = c("0.25", "0.5", "0.75")) 


mpg %>% 
  summarise(mean_displ = mean(displ),
            sd_displ = sd(displ))

mpg %>% 
  group_by(manufacturer) %>% 
  summarise(mean = mean(displ),
            sd = sd(displ)) %>% 
  arrange(desc(mean))
```

# New summarise features - across
```{r}
mpg %>% summarise_if(is.numeric, mean)



mpg %>% summarise(category = c("low", "high"),
                  across(is.numeric, range))

mpg %>% summarise(cyl = mean(cyl),
                  hwy = mean(hwy))

mpg %>% summarise(across(c(cyl, hwy), mean),
                  across(displ, median))

mpg %>% 
  group_by(manufacturer) %>% 
  summarise(n = n(),
            missing = sum(is.na(displ)))


mpg %>% mutate(displ = scale(displ, center = 2, ))
```

