---
title: "200505 Tidy Tuesday"
author: "Vegard Lysne"
date: "5 5 2020"
output: html_document
---

# Packages
```{r setup, include=FALSE}
library(tidytuesdayR)
library(tidyverse)
library(skimr)
library(ggrepel)
```

# Check out the Tidy tuesday package 
```{r}
# List functions of tidy tuesday packages
ls("package:tidytuesdayR")

# List all available data sets
tt_available()

# load a data set using year/week or date
tt <- tt_load(2020, week = 8)

# Look at data
tt

# Load specific data
d <- tt$food_consumption
```

# Explore data set
## Show and summarise data
```{r}
d
str(d)
glimpse(d)
summary(d)
skim(d)
names(d)
```

## visualize variables
```{r fig.height=15}
d %>% 
  ggplot(aes(log(consumption))) +
  geom_density() + 
  facet_wrap(~food_category, scales = "free")

d %>% 
  ggplot(aes(co2_emmission)) +
  geom_density() + 
  facet_wrap(~food_category, scales = "free_y")

d %>% 
  ggplot(aes(log(consumption), co2_emmission)) +
  geom_point() + 
  facet_wrap(~food_category, scales = "free_x")

d %>% 
  group_by(country) %>%
  summarise(CO2 = mean(co2_emmission)) %>% 
  arrange(desc(CO2)) %>% 
  ggplot(aes(x = fct_inorder(country), y = CO2)) + 
  geom_col() + 
  geom_col(data = . %>% filter(country == "Norway"), fill = "red") +
  geom_col(data = . %>% filter(country == "Sweden"), fill = "blue") + 
  coord_flip()
```

```{r fig.height = 15, fig.width = 6}
d %>% 
  filter(country %in% c("Norway", "Sweden", "Denmark", "Finland", "Iceland")) %>% 
  group_by(country, food_category) %>%
  summarise(CO2 = mean(co2_emmission)) %>% 
  ungroup(country, food_category) %>% 
  arrange(desc(CO2)) %>%
  ggplot(aes(y = fct_inorder(food_category), x = CO2)) + 
  geom_col() + 
  facet_grid(rows = vars(country))
```

# Answer some cool questions
## 1. Which country have the highest food-related CO2 emissions per person?
```{r}
d %>% 
  group_by(country) %>% 
  summarise(CO2 = mean(co2_emmission)) %>% 
  arrange(desc(CO2)) %>% 
  slice(1)
```

## 2. Heatmap of top 50 countries and food groups (+ excluding beef)
```{r fig.width = 10, fig.height = 5}
top50 <- d %>% 
  group_by(country) %>% 
  summarise(CO2 = mean(co2_emmission)) %>% 
  arrange(desc(CO2)) %>% 
  slice(1:50) %>% 
  pull(country)

d %>% 
  filter(country %in% top50 & food_category != "Beef") %>% 
  ggplot(aes(y = food_category, x = country, fill = co2_emmission)) + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  labs(y = NULL, x = NULL) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

## 3. Comparison of CO2-emmission from animal food vs non-animal food
```{r}
d %>% 
#  filter(country %in% top50) %>% 
  mutate(catnew = case_when(food_category %in% c("Beef", "Eggs", "Lamb & Goat", 
                                                 "Milk - inc. cheese", "Pork", 
                                                 "Poultry", "Fish")  ~ "Animal", 
                                                 TRUE ~ "Non-animal")) %>% 
  group_by(country, catnew) %>% 
  summarise(CO2 = mean(co2_emmission)) %>% 
#  arrange(desc(CO2)) %>% #(dette rangerte landene etter max CO2 utslipp fra enten animal eller non-animal, ikke totalen)
  mutate(sum = sum(CO2)) %>% arrange(desc(sum)) %>%  # denne linjen ville beregnet summen i hvert land og rangert på totalen
  ggplot(aes(y = fct_inorder(country), x = CO2, fill = catnew)) + 
  geom_col(alpha = 0.4) +
  facet_grid(cols = vars(catnew))
```

## 4. Which countries have higher CO2 from non-animal sources
```{r}
(d %>% 
  mutate(catnew = case_when(food_category %in% c("Beef", "Eggs", "Lamb & Goat", 
                                                 "Milk - inc. cheese", "Pork", 
                                                 "Poultry", "Fish")  ~ "Animal", 
                                                 TRUE ~ "Non-animal")) %>% 
  group_by(country, catnew) %>% 
  summarise(CO2 = mean(co2_emmission)) %>% 
  pivot_wider(names_from = catnew,
              values_from = CO2) %>% 
  filter(`Non-animal` > Animal)) %>% 
  pull(country) -> vegetablecountries
```

# 5. Use the gapminder dataset to plot gdp vs lifeExp for the most current year (2007). Highlight countries with higher CO2 emmission from non-animal foods, and label those with life expectancy > 70 years
```{r}
library(gapminder)

# Siste år kunne vi heller funnet med max(gapminder$year)

gapminder %>% 
  filter(year == 2007) %>% 
  ggplot(aes(gdpPercap, lifeExp)) + 
  geom_point(data = . %>% filter(country %in% vegetablecountries),
             size = 3, color = "red") + 
  geom_point() +
  geom_label_repel(data = . %>% filter(country %in% vegetablecountries & lifeExp > 70), 
                           aes(label = country))
```

