---
title: Volcano eruptions
subtitle: "Tidy tuesday week 20"
author: "Amanda"
date: "11 5 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and data

```{r}

library(tidyverse)
library(tidytuesdayR)
library(ggrepel)

tuesdata <- tidytuesdayR::tt_load(2020, week = 20)
volcano <- tuesdata$volcano
eruptions <- tuesdata$eruptions
events <- tuesdata$events
```

# Which country has the highest number of volcanos?
```{r}
volcano %>% 
  group_by(country) %>%
  summarise(antall = n_distinct(volcano_number)) %>% 
  arrange(desc(antall)) %>%
  slice(1:15) %>% 
  mutate(country = fct_inorder(country) %>% fct_rev()) %>%
  ggplot(aes(country, antall)) +
  geom_col(aes(fill = antall)) +
  scale_fill_viridis_b(option = "inferno") +
  coord_flip() +
  theme_classic()

# color by volcano type
top_15 <- volcano %>% 
  group_by(country) %>%
  summarise(antall = n_distinct(volcano_number)) %>% 
  arrange(desc(antall)) %>%
  slice(1:15) %>% 
  pull(country)
  
volcano %>% 
  filter(country %in% top_15) %>%
  mutate(primary_volcano_type = str_remove(primary_volcano_type, pattern = "\\(.*\\)|\\?")) %>% 
  mutate(country = factor(country, levels = top_15) %>% fct_rev()) %>%
  ggplot(aes(country, fill = primary_volcano_type)) +
  geom_bar() +
  scale_fill_viridis_d(option = "inferno") +
  coord_flip() +
  theme_classic() +
  labs(x = NULL, title = "Number of volcanos", y = NULL, fill = NULL)
```

# Countries with most volcanos that has been active past 50 years
```{r}
volcano %>% 
  left_join(select(eruptions, volcano_number, start_year), by = "volcano_number") %>% 
  filter(start_year > 1970) %>% 
  group_by(country) %>% 
  summarise(antall = n_distinct(volcano_number)) %>% 
  arrange(desc(antall)) %>%
  slice(1:15) %>% 
  mutate(country = fct_inorder(country) %>% fct_rev()) %>%
  ggplot(aes(country, antall)) +
  geom_col(aes(fill = antall)) +
  scale_fill_viridis_b(option = "inferno") +
  coord_flip() +
  theme_classic()
```

# What's the distribution of volcano height within each country?
```{r}
volcano %>% 
  filter(country %in% top_15) %>% 
  ggplot(aes(elevation)) +
  geom_density(aes(y = after_stat(count))) +
  facet_wrap(~country) 
```

# Which countries have volcanos with the most eruptions?
```{r}

```

# What volcano is on average the most explosive? (since year 0)
```{r}

```

# How many different event types are there? What are the most common types of events?
```{r}

```

# Do different event types have different Volcano Explosivity Index?
```{r}
interesting_events <- c("Bombs", "Explosion", "Lava fountains", "Tsunami")

eruptions %>%
  drop_na(vei) %>% 
  left_join(events, by = c("eruption_number", "volcano_number", "volcano_name")) %>% 
  filter(event_type %in% interesting_events) %>% 
  ggplot(aes(x = vei, fill = event_type)) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Volcano Explosivity Index (VEI)", fill = NULL)
```

## Session info
```{r}
devtools::session_info()
```
