---
title: "cheatsheets-purrr"
author: "Jacob J. Christensen"
date: "21 4 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a simple walkthrough of some key functions from [the `purrr` cheatsheet](https://rstudio.com/resources/cheatsheets/). The updated and extended code base can be found at [the tidyverse-purrr website](https://purrr.tidyverse.org/index.html). 

## Load libraries

```{r}
library(purrr) # all `purrr` functions
library(ggplot2) # the "mpg" dataset
library(dplyr) # all `dplyr functions`
```

## The mpg dataset

From `ggplot2`.

```{r}
mpg
```


## Apply functions

### map

What column types do we have in mpg?

```{r}
map(mpg, class)
mpg %>% map(class) # Same, but with pipe (%>%)
res <- mpg %>% map(class)
```

```{r}
class(res)
res$manufacturer
res[["manufacturer"]]
```

What are the unique values of the character variables/columns/vectors?

```{r}
res <- mpg %>% 
  select_if(is.character) %>% 
  map(unique)

mpg %>% 
  map_if(is.character, unique)

res$fl
res %>% pluck("fl")
```


### map_* to simplify

Many variants: map_dbl, map_chr, etc.

```{r}
# Check dimensions in dataset
nrow(mpg)
ncol(mpg)
dim(mpg)

# Check type of single variable
typeof(mpg$manufacturer)

# Check type of all of them, and count them up in a table
mpg %>% 
  map_chr(typeof) %>% 
  table()
```

Use map for summary stats. 

```{r}
res <- mpg %>% 
  select_if(is.numeric) %>% 
  map(mean)

class(res)
res$displ

res <- mpg %>% 
  select_if(is.numeric) %>% 
  map_dbl(mean)

class(res)
typeof(res)
```

Use map inside mutate to map over nested data.

```{r}
library(tidyr)

mpg_nested <- mpg %>% 
  group_nest(cyl)

mpg_nested$data[[1]]
mpg_nested %>% 
  mutate(displ_mean = map_dbl(data, ~mean(.x$displ, na.rm = TRUE)), 
         displ_sd = map_dbl(data, ~sd(.x$displ, na.rm = TRUE))
         )
```


### map2

Use map2 to map over two lists (or columns in a dataset). 

```{r}
mpg_nested
mpg_nested$data[[1]]

res <- mpg_nested %>% 
  mutate(res = map2(.x = cyl, .y = data, 
                    .f = ~paste("Cyl =", .x, .y$manufacturer, .y$model, .y$year)
                    ))

res$res[[3]]
```

### pmap

Use pmap to map over three or more list elements. 

```{r}
res <- mpg %>% 
  group_nest(cyl, drv) %>% 
  mutate(res = pmap_chr(.l = list(cyl = cyl, drv = drv, data = data), 
                    .f = ~paste("Cyl =", ..1, 
                                "Drv =", ..2, 
                                ..3$manufacturer[1], ..3$model[1], ..3$year[1])
                    ))

res
res$res[[1]]
```


### pluck, keep, discard

These are nice to pluck a single list element, or keep/discard all those that meet some logical statement.

```{r}
res <- mpg %>% 
  select_if(is.character) %>% 
  map(unique)

res$trans
pluck(res, "trans")
res %>% pluck("trans")

mpg %>% 
  keep(is.character)

mpg %>% 
  discard(is.character)

res %>% keep(~length(.) > 10)
res %>% keep(function(x) length(x) > 10) # same thing
```


### modify and friends

We didn't go through this, but it's similar to map -- however, it always returns the same type as the input object. 

```{r}
mpg %>% modify_if(is.character, as.factor)

# In this case that would be similar to this mutate call:
mpg %>% mutate_if(is.character, as.factor)
```



### set_names

Very useful to set names to things. 

```{r}
res <- list(1, 2, "this is a list object")

# No names, just numbers [[1]], [[2]] and so on
res

# Set names to it
res %>% 
  set_names(nm = c("jacob", "juel", "christensen")) %>% 
  set_names(toupper)
```



## Modeling example

```{r fig.width=4, fig.height=3}
mpg %>% 
  ggplot(aes(displ)) + 
  geom_histogram()

mpg %>% 
  ggplot(aes(cty, displ)) + 
  geom_point() + 
  geom_smooth(method = "lm")

mpg %>% 
  ggplot(aes(log(hwy), displ)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

Run model. 

```{r}
fit <- lm(displ ~ cty, data = mpg)
fit

fit_tidy <- broom::tidy(fit, conf.int = TRUE)
fit_tidy
```

Run model for each cylinder. 

```{r}
res <- mpg %>%
  group_nest(cyl) %>% 
  mutate(fit = map(data, ~lm(displ ~ cty + hwy, data = .x)), 
         fit_tidy = map(fit, ~broom::tidy(.x, conf.int = TRUE) %>% filter(term == "cty")), 
         estimate = map_dbl(fit_tidy, ~round(.x$estimate, digits = 2)), 
         conf.low = map_dbl(fit_tidy, ~round(.x$conf.low, digits = 2)), 
         conf.high = map_dbl(fit_tidy, ~round(.x$conf.high, digits = 2)), 
         res = paste0("Beta ", estimate, " (95 % CI = ", conf.low, ", ", conf.high, ")")
         )

res
res$fit_tidy[[1]]
res$data[[1]]
```

We could have used factor variables in the model too. 

```{r}
mpg %>% 
  mutate(cyl = factor(cyl)) %>% 
  lm(displ ~ hwy + cyl, data = .) %>% 
  broom::tidy() #%>% 
  # filter(term == "cyl5")
```







```{r}
devtools::session_info()
```


---
